---
title: "Honesty at the Margin"
subtitle: "Experiments on the Dynamics of Deception"
author: "Simon Halliday, Jonathan Lafky and Alistair Wilson"
date: "12/17/2017"
output: ioslides_presentation
---

```{r setup, include = FALSE,  message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(mosaic)
library(plotly)
library(haven)
library(tidyverse)
```


## Context with my work & work by co-authors

- Paper on principal-agent relationships & benefits of abstaining from control accepted at J. Economic Behavior and Organization. 
- Related to this idea of P-A relationships, is a paper I have with a Jonathan Lafky on rating systems (being revised after rejection)
- Jonathan has a related paper with Alistair on incentives for information transmission 

## Context
Evidence in experimental and behavioral economics suggests the following: 

- people have a *preference* for honesty (or against guilt or lying)
    - see Gneezy (2005), Gneezy et al (2013), Dreber & Johannssen (2010), Kartik (2009), and many others
- people have a utility function with standard properties
    - they have *social preferences* because they *dislike* lying 
    - lying imposes a utility cost 
    - **extensive margin**: cost to lying *at all*
    - **intensive margin**: cost in *size* of lie 

## Lying in the real world?
But there's lots of evidence of people lying for profit

- people lied about MBS and CDOs prior to the financial crisis
- financial advisers who are *not* fiduciaries
- medical professionals who benefit from prescribing certain drugs
- lawyers displaying prosecutorial dishonesty
- current political climate?
    - do politicians who lie crowd out those who are unwilling (or are *as* unwilling) to do so?
    - what does this do to the market for politicans? 

## Motivating questions & gaps in literature

- If lies are preference-based, how vulnerable are the preferences to institutional setting? 
    - Do the preferences override the setting, or are the preferences context-dependent? 
- How do people respond to the opportunity to lie when competing with others? 
- Can competition encourage/discourage full deception rather than merely 'disguised' lies (Fischbacher and Föllmi-Heusi, 2013)? 
- How does the size of outside offers affect choices to compete & lie? 
- Do people select into honest settings knowing others will lie? 
- Do people adapt to deception and lie more? (Garrett et al, 2016)

## Design Context

Fischbacher and Föllmi-Heushi (JEEA, 2013) 

- a dice-throwing game to measure *aggregate* deception. 
- subjects roll a 6-sided die and report what they rolled
- Roll of 6 $\Rightarrow$ CHF 0 
- Roll of $1 \rightarrow 5 \Rightarrow$ equivalent CHF
- with large $n$, should be a uniform distribution  $U \sim [0, 5]$ 

## For example...
```{r theoretical_distribution}
# function to generate dice throws
n <- sample(x = 0:5, size = 80, replace = TRUE)
df <- data.frame(n)
ggplot(df, aes(x = factor(n))) +
  geom_bar(aes(y = (..count..)/sum(..count..))) + 
  xlab("Dice Roll") + 
  ylab("Frequency") + 
  geom_hline(yintercept = 1/6) + 
  scale_x_discrete(labels = c("0", "1", "2", "3", "4", "5")) +
  theme_bw()
  ggplotly()
```


## FFH's Results
```{r ffh_results}
FFH <- read_stata("more/ffh2013/LiesInDisguiseRepeated.dta")
FFH %>% 
  filter(!is.na(Participation1)) %>% 
  ggplot(aes(x = factor(Participation1))) + 
  geom_bar(aes(y = (..count..)/sum(..count..))) + 
  xlab("Dice Roll") + 
  ylab("Frequency") + 
  geom_hline(yintercept = 1/6) + 
  annotate("text", label = "theoretical ", x = 2, y = 0.25, size = 5) +
  annotate("text", label = "distribution", x = 2, y = 0.22, size = 5) +
  geom_segment(aes(x = 2, xend = 2, y = 0.21, yend = 0.17), arrow =
                 arrow(length = unit(0.25, "cm"), type = "closed")) +
  scale_x_discrete(labels = c("0", "1", "2", "3", "4", "5")) +
  annotate("text", label = "lies in ", x = 5, y = 0.38, size = 5, colour = "red") +
  annotate("text", label = "disguise", x = 5, y = 0.36, size = 5, colour = "red") + 
  geom_segment(aes(x = 5, xend = 5, y = 0.34, yend = 0.28), arrow = arrow(length = unit(0.25, "cm"), type = "closed"), col = "red") +
  theme_bw()
  ggplotly()
```


## FFH Continued 

-  also did a ‘high-stakes’ treatment 
    - subjects were paid CHF3 per pip reported up to CHF15
- choices all one-shot and the majority of the analysis is between subjects 
- Because the experiment was run after other experiments (as a 10-minute add-on), some subjects participated more than once. 
- Subjects who participated more than once were more likely to report a higher number the second time. 
    - do we 'learn to lie'?

## Extension: D & K (2017)

Dannenberg and Khachatryan (2017, working paper) 

 - replicate FFH + competition + group decisions
 - one-shot competitive choice
    - either with one other participant, or 
    - as a member of a team of three against another team of three, or
    - as a single or team of three against a single or team of three (2 by 2 treatment design). 
 - subjects who compete one-on-one are *no more dishonest* than subjects who do not compete when making their choices (between subjects)
 - teams of three against teams of three are more likely to report higher numbers (are more dishonest) 
 - consistent with Charness & Sutter (2013) on groups making "more self-interested" decisions 


## Gaps left by FFH, D & K?

- Subjects in a "competitive" treatment have no choice about competing
- Group vs. individual not that interesting to us 
- D & K have no opportunity to "learn to lie"
- If a subject cannot choose *not* to compete, is it really a *choice to compete*? 
- What is the role of an outside offer/opportunity cost? 
- What about the margins of: 
    - choosing to compete and
    - reports of those who compete conditional on having chosen to compete? 

## Our design (HLW)

T1: Baseline Treatment (Decision problem)

- Choice of two options, A or B. 
-  If A is chosen, subject reports a 10-sided die roll, which is compared to three random numbers generated by a bingo cage (automated). 
    - If the report matches any selected number, they receive $20.
    - Otherwise payment is $0.00.
- If B is chosen, the subject reports a 10-sided die roll. 
    - Roll compared with a publicly rolled 10-side die. 
    - If higher, player obtains $20; 50-50 chance on a tie. 

## Our design (HLW, contd.)

T2: Treament "Automated High Roll" (Decision problem)

- Same as T1, except zero probability of winning if report is 0 through 8
- 50-50 chance of prize if reporting 9.
- Why? Upper-bound limit on those who want the 50-50 lottery if they are guaranteed to lie. 

## Our design (HLW, contd.)
T3: Competition (Strategic)

- Each session has four (?) groups of 6 (?) subjects
- Choice of two options, A or B. 
-  If A is chosen, subject reports a 10-sided die roll, which is compared to three random numbers generated by a bingo cage (automated). 
    - If the report matches any selected number, they receive $20.
    - Otherwise payment is \$0.00.
- If B is chosen, the subject reports a 10-sided die roll. 
    - Roll compared with a another player's reported 10-side die roll. 
    - If higher, player obtains \$20; 50-50 chance on a tie. 


## Our design (HLW, contd.)
Additional details 

- Subject repeats their choice 24/25 times. 
- We observe the margin of choosing Options A or B (extensive)
- On choosing B, we measure the distribution of reports (intensive because repeated within subject)
- Contrast with theoretical distribution and across treatments

## Potential behaviors (randomized n = 100 rolls)
```{r hlw_distribution}
# function to generate dice throws
d10 <- sample(x = 0:9, size = 100, replace = TRUE)
hlw <- data.frame(d10)
ggplot(hlw, aes(x = factor(d10))) +
  geom_bar(aes(y = (..count..)/sum(..count..))) + 
  xlab("Dice Roll") + 
  ylab("Frequency") + 
  geom_hline(yintercept = 1/10) + 
  scale_x_discrete(labels = seq(0, 9, 1)) +
  theme_bw()
  ggplotly()
```


## Conclusions and contributions?

- Measure margins not previously measured
- Choices are repeated, so no concern that subjects don't know what they're doing (others so far are one-shot). 
- Task is relatively short with a high expected payoff for the short task (incentive compatible)
- We're still decided on how many of the 24/25 choices to pay (2 or 3?)
- Extensions
    - experiment to examine demand and supply of deception
    - what happens if hiring decisions are based on report? $\uparrow$ or  $\downarrow$ deception? 
    
