---
title: "Simple plotting exercise"
output:
  html_document: default
  html_notebook: default
---

Here is some code to make some basic plots and differentiate them up by a categorical variable (also called a **factor** variable in R).

First, we will clear the memory and load in the `tidyverse` package, which we will need for plotting. The `tidyverse` is a covenient collection of the packages we will need to tidy and plot our data. We read in a data file, in this case a comma separate value file (`csv`).

```{r echo = FALSE, message = FALSE}
rm(list = ls())
library(tidyverse)
```

Now, we will load in a dataset. 

The data used are from Hart, O'Hara, and Sirianni (2017)
"The Effect of Sustainability Commitments on College Admissions: Evidence from the Presidents’ Climate Commitment"
A brilliant paper that is currently being drooled over by referees at *JEEM*.
```{r echo = FALSE, message = FALSE}
sat.by.type <- read_csv("GroupedbyType.csv")
```


This data has two time series in it, showing the average 75th percentile SAT score for each year for each type of school, public vs private. We use `ggplot()`, a package developed by Hadley Wickham, to plot our data. The `gg` in `ggplot` stands for Grammar of Graphics, which we employ to specify the different characteristics of our graphs. 

We start by making a simple scatterplot of the points in the data. 
```{r echo = FALSE}
ggplot(sat.by.type, aes(x = year, y = avg.sat)) + 
   geom_point() + 
  labs(title = "SAT scores (75th percentile) by school type", y = "SAT")
```

The first line in the code performs a variety of functions. It starts by calling on `ggplot()` and tells it that we are going to use the dataframe (an object we have called `sat.by.type`). 

The dataframe is the first argument of the `ggplot()` command. 

The second argument is an argument `aes()` for *aesthetics*. What are aesthetics? The aesthetics of the graph:

- Specify the variables on the $x$ and $y$ axes
- Specify whether to attribute *colors* to different characteristics, e.g. later saying `color = SchoolType`
- Specify whether to attribute different *fills* (e.g. filling different columns/bars with different colors) to different categories, e.g. `fill = SchoolType` (if we had bars or other types of plots)
- Specify whether to attribite different *shapes* to different categories, e.g. public schools given stars and private schools triangles
- Notice that editing the `aes()` function in the original `ggplot()` call results in changes to *all* subsequent graph properties. This is *good* because it can make change a lot of properties quick, but *bad* if we only want to edit the properties of one layer of the graph. 

But just calling on `ggplot()` doesn't result in an actual graph we can interpret until we tell it what *kind* of plot we want. In the case above, we wanted a scatterplot. 

`ggplot()` operates by using *layers* to add different aspects to graphs (similar to what you do if you edit a photo in PhotoShop).
Layers incorporate the following ideas: 

- a layer of dots, lines, bars to produce graphs 
- a layer for a graph title, axis labels 
- annotations to the different points on the graphs, e.g. adding $R^2 = 0.92$ to a particular coordinate in a figure 
- editing the legend for the graph 

To add a layer to our plots, we need to end the *preceding* line with a `+` sign to tell `ggplot()` that it's job isn't done yet. So our code will look like the following: `ggplot(sat.by.type, aes(x = year, y = avg.sat)) + ` (the plus sign indicating something is about to come up). If you end a line of code with a plus sign, then R (and `ggplot()`) expect you to put something there, so if you put that into your *console* it'll just sit there waiting for you to add something or to hit [esc] to break the process. 

The different kinds of graphs are called `geom_`s in ggplot parlance. A scatterplot requires us to use `geom_point()`. Notice that because we have *already* specified what the x and y variables are, we do *not* need to do so again in the `geom_point()` function Later, though, we will add attributes in the `geom_point()` function to specify particular attributes of the scatter plot that we want to define. 

We can give the points different colors by school type. We need to edit the 
```{r echo = FALSE}
###  Make a simple scatterplot of average SAT scores
###  and color points by variable SchoolType
ggplot(sat.by.type, aes(x = year, y = avg.sat)) + 
   geom_point(aes(color = SchoolType)) + 
  labs(title = "SAT scores (75th percentile) by school type", y = "SAT")
```

Suppose we want to make a simple line graph to compare the trends of SAT scores over time in the different types of schools. 

```{r echo = FALSE}
###  Make a line plot of average SAT scores by school type
ggplot(sat.by.type, aes(x = year, y = avg.sat, group = SchoolType)) + 
   geom_point(aes(color = SchoolType)) + 
   geom_line() +
  labs(title = "SAT scores (75th percentile) by school type", y = "SAT")
```
But what's happening above?

- We've added `geom_line()` to add lines to our time series. 
- But the lines are black because we didn't specify a color aesthetic in the `geom_line()` function. 
- There are two ways to get around this
  - add `aes(color = SchoolType)` to the `geom_line()` function (OK if we only want color to apply to the lines)
  - add `color = SchoolType` to the `aes()` call in the original `ggplot()` call (if we want the color grouping to apply to *all* subsequent geoms - so often more efficient if we want to change an aspect of all aesthetics of our graph)

```{r}
###  Make a line plot of average SAT scores by school type
ggplot(sat.by.type, aes(x = year, y = avg.sat, group = SchoolType, color = SchoolType)) + 
   geom_point() + 
   geom_line() +
  labs(title = "SAT scores (75th percentile) by school type", y = "SAT") + 
  scale_color_discrete(name = "Type of School", 
                       breaks = c("private", "public"), 
                       labels = c("Private institution", "Public Institution"))
```








