---
title: "Lab 5: Dummies, Interactions and Demand for Fish"
output:
  html_document:
    css: ../stylesheets/lab.css
    highlight: pygments
    theme: cerulean
bibliography: ../more/fish_bib.bib
biblio-style: chicago 
---

## Introduction

In economics, there is a strong history trying to understand market behavior. 
Properly identifying a market demand function is hard, but various authors have studied fish markets as examples of markets where a relatively homogeneous good (fish) with a short lifespan (its quality deteriorates if not bought promptly) is sold in competitive regional markets, such as the fish markets at Saumaty in Marseille, France or MERITAN in Ancona, Italy. 
Alan Kirman has been involved in a lot of this work in the past few decades trying to understand the complex interactions that produce market demand functions from not particularly consisten individual demand functions. 
@hardle95 was one of his first papers to study the Marseille fish market, which he returned to for his book [@kirman2010]. He has also done a variety of work with co-authors on fish markets, such as that on the Ancona fish market [@gallegati12]. 
Prof Kirman has kindly provided me with the data from his Marseille studies and so we have the opportunity to look at this rather marvellous data set. 

- Go here on Moodle to download the data: [https://moodle.smith.edu/mod/resource/view.php?id=405319](https://moodle.smith.edu/mod/resource/view.php?id=405319)
- The codebook to understand what the data means is available on Moodle here: [https://moodle.smith.edu/mod/resource/view.php?id=405318](https://moodle.smith.edu/mod/resource/view.php?id=405318)  

```{r requirements, message = FALSE, warning = FALSE}
library(tidyverse)
```


Import the data as follows. 
Notice that this is *different* to how we've historically imported data. This is because the data are in a .txt file and so we can't use `read.csv()`. Instead, we use `read_table()` from the `readr` package in the `tidyverse`. 
```{r import-Marseille}
MarseilleBase <- read_table("../more/MARS.txt", col_names = FALSE)
```

Look at the .txt file. Why do you think we had to specify the option `col_names = FALSE`? 

`head()` the data to look at it. It has an issue in one column (find the issue)
```{r check-Marseille}
head(MarseilleBase)
```

As you can see, we need to solve this problem. We can do this using the `separate()` function in `dplyr`, which is part of the `tidyverse`.
```{r separation}
Marseille <- 
  MarseilleBase %>% 
  separate(X3, c("X3a", "X3b"), sep = 14)
head(Marseille)
```

What did I do above? 

- Create a new data frame called `Marseille`
- Using the data frame `MarseilleBase`
- I use `separate` on the variable `X3`
- I ask `separate` to create two new variables: `X3a` and `X3b`
- Because the txt file has the names of the buyers limited to 14 characters, I ask `separate` to separate the variables at the 14th character
- I can now work with this data frame and rename my variables sensibly. 

```{r renaming}
Marseille <- 
  Marseille %>%
  rename(sellerb = X1, 
         lot = X2, 
         nbuyer = X3a,
         seller = X3b,
         date = X4, 
         buyerc = X5, 
         fish = X6, 
         tempa = X7, 
         tempb = X8) 
Mars <- 
  Marseille %>%
  mutate(weight = tempa/10, 
         price = tempb/100, 
         fish = parse_number(fish))
head(Mars)
```

- DO NOT run the code below yourselves. 
- The sample size is huge and we don't want to overload the server. 
- Just look at the plot. What does it suggest?  
- What does the line correspond to? 
- What kind of pattern are you able to see? 

```{r AllFishPlotLin, warning = FALSE, message = FALSE}
Mars %>% 
  ggplot(aes(x = weight, y = price)) + 
  geom_point() + 
  stat_smooth(method = "lm", formula = y ~ x,  size = 1) +
  ylim(0, 100)
```


Note, we don't need a *linear* downward-sloping function...

- (Again DO NOT RUN THIS CODE)

```{r AllFishPlotNonlin, warning = FALSE, message = FALSE}
Mars %>% 
  ggplot(aes(x = weight, y = price)) + 
  geom_point() + 
  stat_smooth(method = "auto", size = 1) +
  ylim(0, 100)
```



## A basic linear regression
Notice that you can run a linear regression with the whole sample of fish: 
```{r linear-model1, warning = FALSE, message = FALSE}
m1 <- lm(price ~ weight, data = Mars)
summary(m1)
```

- What is the sign on the coefficient for weight? 
- What is weight's significance? 
- What does the relationship mean in terms of economic theory? 

## Filtering the Fish
Consult the Marseille Data Description document to check what the different kinds of fish are. 

I have generated  the following graph for one of the kinds of fish (Merlan) by filtering and then graphing the data. 

Filtering for Merlan:
```{r MarlanData, warning = FALSE, message = FALSE}
MarsFish44 <- 
  Mars %>%
  filter(fish == 44)
```

Graphing the Merlan data:
```{r MerlanPlot, warning = FALSE, message = FALSE}
MarsFish44 %>% 
  ggplot(aes(x = weight, y = price)) + 
  geom_point() + 
  stat_smooth(method = "lm", formula = y ~ x, size = 1) +
  ylim(0, 100)
```

We can also run a linear regression just with the filtered Merlan data. 
```{r MerlanRegressions, warning = FALSE, message = FALSE}
OLSMerlan <- lm(price ~ weight, data = MarsFish44)
summary(OLSMerlan)
```

Do the following: 

- Choose a fish in which you are interested and construct a data frame that contains only that kind of fish.
- Construct a plot for that fish's quantity-price observations
- Run a linear regression model of that fish's price-quantity relationship
- Summarize the regression output. 
- Submit your html of this later. 


# Multiple regression with dummy variables
We want to include dummy variables for the different kinds of fish in our regressions. So we need to check whether the fish variable is numeric or a factor. We could `head()` the data and see, but we can also get a "TRUE" or "FALSE" report using the `is.numeric()` or `is.factor()` functions. 
```{r}
is.numeric(Mars$fish) #This must be true because I made it numeric
is.factor(Mars$fish) #This must be false unless you change it
```

As you can see, we need to change `fish` into a *factor* variable because we want to use the factors as dummy variables. 

- How would you do this? 
- Do the variables have the labels you'd like them to have? (i.e. the names of the fish)
- Is there a way to remedy this? 

### Some ideas for factors
We need to take the information from the Marseille Data Details document and create two sets of columns using the following idea: `ObjectName <- c(content1, content2, ..., contentn)`. 

I would recommend you call the one object (with the numbers) `FishLevels` because it corresponds to the *levels* of your factor variable. 

I would recommend you call the second object `FishLabels` because it corresponds to the *labels* of the fish that you want. 

You will then use the function `factor()` inside the function `mutate()` (which you already know). 

You would then do something like the following (note this won't work because I don't have actual values attributed to these objects yet).
```{r, eval = FALSE}
MarsFactorFish <- 
  Mars %>% 
  mutate(fishvariablename = factor(fishvariablename, 
                                   levels = FishLevels, 
                                   labels = FishLabels))
```


You should get something like the following when you head your data: 
```{r re_factor, echo = TRUE}
FishLabels <- c("Merlan", "Carrelet", "Cabillaud", "Loup/Bar", " Raie Ange", "Raie (Entiere)", "Sole", "Turbot", "Cabillaud (Tranches)", "Sole (Filet)", "Annarrhyque (Filets)", "Loup (Filets)")
FishLevels <- c(44, 62, 77, 122, 204, 205, 214, 244, 285, 289, 314, 315)
Mars <- 
  Mars %>%
  mutate(fish = factor(fish, 
                       levels = FishLevels, 
                       labels = FishLabels))
head(Mars)
```

## Fish Factors in the Multiple Regression
Now, include the fish variables as factors in our regression 
```{r model_dummies}
m2 <- lm(price ~ weight + fish, data = Mars)
summary(m2)
```

- What do the different coefficients on the categorical (dummy) variables mean? 
- Can you provide new output for the fish regressions using stargazer? 


```{r model_interaction}
m3 <- lm(price ~ weight + fish + weight*fish, data = Mars)
summary(m3)
```

- What does the model say? (that is, interpret the *equation* of the regressionl line)
- What is an *interaction term*?
- What do the coefficients on each of the variables mean now?

## References