---
title: 'Lab Extras: Ifelse (Using Sutter, 2009)'
author: "Simon Halliday"
date: "Spring 2019"
output:
  html_document:
    css: ../stylesheets/lab.css
    highlight: pygments
    theme: cerulean
#bibliography: skeleton.bib
link-citations: yes
---


## Loading required packages

Be sure to load the packages you require.
We use the "Tidyverse" of packages. 

```{r requirements, message = FALSE, warning=FALSE}
library(tidyverse)
```



## Data 
I'm going to re-import the Sutter dataset we used in Lab 4. 
```{r read_csv, message=FALSE, warning = FALSE}
SutExp <- read_csv("../more/sutterexperiment.csv")
```
If you don't have the data accessible, you can download it here: [https://drive.google.com/file/d/0B9jjwkjdUJU7cElvcVI5U3hhdEU/view?usp=sharing](https://drive.google.com/file/d/0B9jjwkjdUJU7cElvcVI5U3hhdEU/view?usp=sharing). 


## Using rename to have consistent variable names
I'm also going to re-run the code `tolower` for consistency. 
```{r}
names(SutExp) <- tolower(names(SutExp))
```

## Narrow the data 
We're now going to narrow the data to get a long/narrow dataset with rounds and values in columns. 
```{r narrowing}
NarSut <- 
  SutExp %>% 
  gather(round, value, -treatment, -team, -subject, -session) %>% 
  arrange(session, treatment, subject)
```


## Using `ifelse`
Imagine you wanted to aggregate (`group_by`) on a variable that indicated "before" and "after" the first five rounds. 
If we wanted to do this, one of the easiest ways to do so would be first to create an indicator variabe that takes the value 1 for rounds 1 to 5 and 0 otherwise. 
To do so, we're going to use `mutate()` to create a new variable and `ifelse()` to make the creation of that variable conditional on the values of other variables. 
How does `ifsels()` work? It is like a question that asks "If a condition is true, then report a value I tell you to, if false report another value I tell you to."
So here, we want to say the following 

- If the round is round 1, round 2, round 3, round 4, or round 5 
- THEN the value if the condition is TRUE is 1
- OR the value if the condition is FALSE is 0

```{r ifelse1}
NarSut <- 
  NarSut %>% 
  mutate(early = ifelse(round %in% c("r1", "r2", "r3", "r4", "r5"), 1, 0)
         )
```

You can also embed `ifelse()` functions in each other. For example, imagine I wanted a variable with three values, "beginning", "middle", and "end", then I could do the following. 
```{r ifelse2}
NarSut <- 
  NarSut %>% 
  mutate(stage = ifelse(round %in% c("r1", "r2", "r3"), "beginning",
                        ifelse(round %in% c("r4", "r5", "r6"), "middle", "end"
                               ) #Notice this line break
                        ) #And this one. 
         )
```
What function is played by the line breaks in the code? 

You can also use `ifelse()` with other variables as the outcomes rather than values or characters as I've done above. Before showing you that I'm going to manipulate the data a bit. 

First generate a unique ID: 
```{r uniques}
NarSut <- 
  NarSut %>%
  mutate(uniqid = paste(session, treatment, subject, sep = "_"))
```

Second, create a variable that is the maximum reported value someone gives and I'm creating a variable that is the median of the sample for all values. (Note: why is it important for me to use the function `ungroup()` in the code chunk?) 
```{r personalmax}
NarSut <- 
  NarSut %>%
  group_by(uniqid) %>% 
  mutate(maxval = max(value)) %>% 
  ungroup() %>% 
  mutate(medval = median(value))
```

Third, I'm going to create a new variable which reports the `maxval` if `maxval > medval` and  reports `medval` otherwise. 
```{r ifelse3}
NarSut <- 
  NarSut %>%
  mutate(maxormed = ifelse(maxval > medval, maxval, medval
                           )
         )
```
I could now use the variable `maxormed` in calculations if we wanted to. Because reasons. 

To review, what have we done with ifelse? 

- We've created variables that met a condition, and reported a number 1 or 0 
- We've created variables that met a conditoin, and reported "beginning", "middle", or "end" (characters or factors if we made them such)
- We've created  variables that met a condition and reported  other variables as results. 

