### Text

Text can decorated with **bold** or *italics*.  It is also possible to 

- create [links](http://rmarkdown.rstudio.com/) 
- include mathematics like $e=mc^2$ or 
$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2$$
- embed images 

Be sure to put a space after the - when you are creating bullets and a space after # when 
creating section headers, but **not** between $ and the mathematical formulas.

### Graphics

If the code of an R chunk produces a plot, this plot can be displayed in the resulting file.

```{r}
source("http://www.openintro.org/stat/data/arbuthnot.R")
plot(arbuthnot$girls)
```

### R output
Other forms of R output are also displayed as they are produced.

```{r}
head(arbuthnot$girls)
```

You can also use "chunk options" to modify how your R chunks work.  We can hide the code:

```{r, echo = FALSE}
summary(arbuthnot$girls)
```

Or we can hide the output:

```{r, eval=FALSE}
summary(arbuthnot$girls)
```

### In-text Code

In a similar vein, you can refer to R code in a paragraph by setting it off with single
ticks, say if we wanted to mentioned that we had used the `summary()` function.  You can
also put R output directly into text, as in: the mean number of births was 
`r mean(arbuthnot$girls)`.

### Destination formats

This file can be knit to HTML, PDF, or Word.  In RStudio, just select the desired 
output file type and click on `Knit HTML`, `Knit PDF`, or `Knit Word`.  Use the 
dropdown menu next to that to change the desired file type.



### Stress Test
Please do not run the following code until the very end of class.

```{r eval=FALSE, echo=FALSE}
set.seed(485)
m <- matrix(rnorm(4e6), ncol = 2000)
system.time(m_inv <- solve(m))
```

###Note
This was gleefully pilfered from Andrew P. Bray (with his permission)


